---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Getting data


Dates of viper observations in Stockholms län 2019-2021 downloaded from artportalen
```{r}
library(tidyverse)
obs_dates <- read_csv("huggorm.csv")
obs_dates
```

Summarise as number of observations per date
```{r}
daily_obs <- count(obs_dates, date, name = "n_vipers") 
daily_obs
```

Daily max temperature at Bromma airport downloaded from SMHI's API
```{r}
max_temp <- read_csv2("https://opendata-download-metobs.smhi.se/api/version/1.0/parameter/20/station/97200/period/corrected-archive/data.csv", skip = 8) %>% 
  select(date = `Representativt dygn`, max_temp = Lufttemperatur) %>% 
  filter(lubridate::year(date) %in% 2019:2021) %>% 
  mutate(max_temp = as.numeric(max_temp)) # Daily 
max_temp
```


Hourly relative humidity (%) downloaded from SMHI's API and summarised as daily average over 10:00-18:00
```{r}
humidity <- read_csv2("https://opendata-download-metobs.smhi.se/api/version/1.0/parameter/6/station/97200/period/corrected-archive/data.csv", skip = 8) %>% 
  select(date = Datum, time = `Tid (UTC)`, humidity = `Relativ Luftfuktighet`) %>% 
  filter(lubridate::year(date) %in% 2019:2021, 
         lubridate::hour(time) %in% 10:18) %>% 
  group_by(date) %>% 
  summarise(humidity = mean(humidity))
humidity
```

Join tables and add flag for weekend, day of year and flag for "wet" weather (more than 60% relative humidity)
```{r}
weather_data <- full_join(max_temp, humidity, by = "date")
weather_data

full_data <- left_join(weather_data, daily_obs, by = "date") %>% 
  mutate(n_vipers = ifelse(is.na(n_vipers), 0, n_vipers), # Turn NAs to zero
         weekend = ifelse(lubridate::wday(date) %in% c(1, 7), 1, 0), # Day 1=Sunday, 7=Saturday
         day = lubridate::yday(date), # Julian day
         wet = ifelse(humidity > 60, 1, 0),
         ) 
full_data
```

## Exploratory 

More vipers on weekends:
```{r}
full_data %>% group_by(weekend) %>% 
  summarise(mean(n_vipers))
```


More vipers on dry days (but winter months are wet!):
```{r}
full_data %>% group_by(wet) %>% 
  summarise(mean(n_vipers))
```

Smoothed daily averages, high-season in March-May:
```{r}
full_data %>% ggplot(aes(x = date, y = n_vipers)) + 
  geom_point(alpha = .3) + 
  geom_smooth(method.args = list(family = "poisson"), method = "gam", formula = y ~ s(x, k = 15)) +
  theme_bw()
```

High-season, less vipers on cold days:
```{r}
full_data %>% filter(lubridate::month(date) %in% 3:5) %>% 
  ggplot(aes(x = max_temp, y = n_vipers)) + 
  geom_point(alpha = .3) + 
  geom_smooth(method.args = list(family = "poisson"), method = "gam") +
  theme_bw()
```

High-season, less vipers on wet days:
```{r}
full_data %>% filter(lubridate::month(date) %in% 3:5) %>% 
  ggplot(aes(x = humidity, y = n_vipers)) + 
  geom_point(alpha = .3) + 
  geom_smooth(method.args = list(family = "poisson"), method = "gam") +
  theme_bw()
```

# Model fitting

This is tricky due to the complex interactions between weather, season, people and vipers. Needs a clear goal. Simple example using a Poisson regression model with smooth terms for day and max_temp (restricted to high-season). 


```{r}
library(mgcv)
model_fit <- gam(n_vipers ~  s(max_temp, k = 10) + s(day, k = 10) + wet + weekend, 
                 family = "poisson", 
                 data = full_data %>% filter(lubridate::month(date) %in% 3:5))
summary(model_fit)
```

Plots of smooth terms (max_temp and day)
```{r}
plot(model_fit, pages = 1, shade = TRUE, trans = exp)
```




